---
title: "Introduction to dtametaTMB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dtametaTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette introduces the main functionality of the
`dtametaTMB` package for meta-analysis of diagnostic test
accuracy (DTA) studies. We demonstrate how to fit three
commonly used models:

- the Reitsma bivariate model
- the Rutter and Gatsonis (HSROC) model
- the Hoyer model for multiple thresholds

using real datasets included in the package.

## How do I load the package?

```{r setup}
#| echo: TRUE
library(dtametaTMB)
```


## Which model should I use?

The Reitsma and Rutter-and-Gatsonis models are appropriate when a single threshold per study is reported. 

  -   The Reitsma model is most appropriate when the studies report true negatives, true positives, false negatives and false positives using the same or similar thresholds, 
  -   whereas the Rutter and Gatsonis model is most appropriate when the studies use different thresholds. 

The Hoyer model allows for multiple thresholds within studies.


## How do I perform a diagnostic test meta-analysis using the bivariate binomial-model of Reitsma?

Unlike the mada package, dtametaTMB uses the binomial likelihood to model the within study distribution as recommended by The Cochrane Handbook for Systematic Reviews of Diagnostic Test Accuracy.

For sensitivity in study $i$, we have the number of diseased individuals testing positive: $y_{Ai} \sim \mathcal{B}(n_{Ai},\pi_{Ai})$.

Similarly for specificity, we have the number of non-diseased individuals testing negative: $y_{Bi} \sim \mathcal{B}(n_{Bi},\pi_{Bi})$.

At the study level, we have

$$ 
\begin{pmatrix}
\operatorname{logit}(\pi_{Ai}) \\
\operatorname{logit}(\pi_{Bi})
\end{pmatrix}
=
\begin{pmatrix}
\mu_{Ai} \\
\mu_{Bi}
\end{pmatrix}
\sim \mathcal{N}
\left(
\begin{pmatrix}
\mu_A \\
\mu_B
\end{pmatrix},
\;
\Sigma
\right),
\quad \text{with} \quad
\Sigma =
\begin{pmatrix}
\sigma_A^2 & \sigma_{AB} \\
\sigma_{AB} & \sigma_B^2
\end{pmatrix}.$$

```{r}
#| echo: TRUE
data("anticcp")
reitsma <- fitReitsma(data=anticcp,
                      TP=TP,
                      FP=FP,
                      FN=FN,
                      TN=TN,
                      study=study)
reitsma
summary(reitsma)
```

## How do I get a summary plot of the Reitsma model?

```{r}
#| fig.height: 8
#| fig.width: 8
#| echo: TRUE
plot(reitsma)
plot(reitsma,scale=0.01)
```

## How do I get a coupled forest plot?

```{r}
#| fig.height: 11
#| fig.width: 13
#| echo: TRUE
forest(reitsma)
```

## How do I perform a diagnostic test meta-analysis using the Rutter and Gatsonis model?

The number of diseased individuals from study $i$ who test positive is denoted by $y_{i1} \sim \mathcal{B}(n_{i1},\pi_{i1})$.

Similarly, the number of non-diseased individuals who test positive is $y_{i2} \sim \mathcal{B}(n_{i2},\pi_{i2})$.

At the study level, we have

$$\operatorname{logit}(\pi_{ij}) =
\left( \theta_i + \alpha_i x_{ij} \right)
\exp(-\beta x_{ij}), \quad \quad \quad
\alpha_i \sim \mathcal{N}(\Lambda, \sigma_\alpha^2),
\qquad
\theta_i \sim \mathcal{N}(\Theta, \sigma_\theta^2),$$

where $$x_{ij} =
\begin{cases}
-0.5 & \text{for non-diseased individuals}, \\
\phantom{-}0.5 & \text{for diseased individuals}.
\end{cases}$$

```{r}
#| echo: TRUE
#| results: 'hide'
data("RF")
ruttergatsonis <- fitRutterGatsonis(data=RF,
                                    TP=TP,
                                    FP=FP,
                                    FN=FN,
                                    TN=TN,
                                    study=study)
```

```{r}
#| echo: TRUE
ruttergatsonis
summary(ruttergatsonis)
```

## How do I get a summary plot of the Rutter and Gatsonis model?

```{r}
#| fig.height: 8
#| fig.width: 8
#| echo: TRUE
plot(ruttergatsonis)
plot(ruttergatsonis,size="se",scale=0.01)
```

## How do I get a coupled forest plot?

```{r}
#| fig.height: 14
#| fig.width: 13
#| echo: TRUE
forest(ruttergatsonis)
```

## How do I perform a diagnostic test accuracy meta-analysis with multiple thresholds per study using the Hoyer method?

The Hoyer model assumes that the diagnostic test measurement arises from a latent continuous variable $Y_{j}$, with $j=0$ for the non-diseased individuals and $j=1$ for the diseased, which follows a parametric survival distribution. The observed test results correspond to interval-censored observations of this latent variable due to the use of thresholds.

### Which distributions does `dtametaTMB` support?

The latent variable $Y_{j}$ is assumed to follow a location–scale family:

$$Y_{j} \sim F(\mu_{j}, \phi_j),$$

where:

- $\mu_j$ is a study-specific location parameter
- $\phi_j$ is a scale parameter

Common choices for $F$ include:

#### Weibull distribution

$$f(y; \mu_j, \phi_j) = \mu_j \phi_j y^{\phi_j - 1} \exp(-\mu_j y^{\phi_j})$$

#### Log-normal distribution

$$f(y; \mu_j, \phi_j) =
\frac{1}{\phi_j y}
\phi\!\left(\frac{\log(y) - \mu_j}{\phi_j}\right)$$

#### Log-logistic distribution

$$f(y; \mu_j, \phi_j) =
\frac{\mu_j \phi_j y^{\phi_j - 1}}{\left(1 + \mu_j y^{\phi_j}\right)^2}$$

### Random effects structure

To account for between-study heterogeneity and the correlation between diseased and non-diseased groups, we introduce study-specific random effects:

$$\log(y_{0}) = \beta_{0} + \varepsilon_0 + u_{0i}, \quad
\log(y_{1}) = \beta_{1} + \varepsilon_1 + u_{1i},$$

with

$$\begin{pmatrix}
u_{0i} \\
u_{1i}
\end{pmatrix}
\sim
\mathcal{N}
\left(
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\begin{pmatrix}
\sigma_0^2 & \rho \sigma_0 \sigma_1 \\
\rho \sigma_0 \sigma_1 & \sigma_1^2
\end{pmatrix}
\right).$$

## Censoring mechanism

Because only thresholded outcomes (TN,TP,FN,FP) are observed, the latent variable $Y_{j}$ is subject to censoring.

- Left censoring: $(-\infty,y_k^R]$
- Interval censoring between adjacent thresholds
- Right censoring: $(y_k^L, +\infty)$

where $y_k^L$ and $y_k^R$ denote finite threshold values. 

### How do I compute sensitivities and specificities?

We can model the ROC curve by predicting sensitivities and specificities (1 - false positive rates) at several threshold values $y$.

#### Weibull distribution
$$S_1(y) = \exp(-(y\,\exp(-\beta_1))^{1/\lambda_1}), \quad \text{with} \quad \beta_1=-\log{(\mu_1}), \quad \lambda_1 = 1/\phi_1,$$
$$S_0(y) = \exp(-(y\,\exp(-\beta_0))^{1/\lambda_0}), \quad \text{with} \quad \beta_0=-\log{(\mu_0}), \quad \lambda_0 = 1/\phi_0.$$

#### Log-normal distribution
$$
S_1(y) = 1-\Phi\left(\frac{1}{\lambda_1} ( \log(y) - \beta_1 )  \right), \quad \text{with} \quad \beta_1=\mu_1, \quad \lambda_1 = \phi_1,
$$
$$
S_0(y) = 1-\Phi\left(\frac{1}{\lambda_0} ( \log(y) - \beta_0 )  \right), \quad \text{with} \quad \beta_0=\mu_0, \quad \lambda_0 = \phi_0.
$$

#### Log-logistic distribution

$$S_1(y) = 1/\left(1 + \exp(-\beta_1/\lambda_1 )\,y^{1/\lambda_1} \right), \quad \text{with} \quad \beta_1=-\phi^{-1}\log{(\mu_1}), \quad \lambda_1 = 1/\phi_1,$$
$$S_0(y) = 1/\left(1 + \exp(-\beta_0/\lambda_0 )\,y^{1/\lambda_0} \right), \quad \text{with} \quad \beta_0=-\phi^{-1}\log{(\mu_0}), \quad \lambda_0 = 1/\phi_0.$$

#### How do we interpret test direction?

For tests where larger values indicate disease (`testdirection = "greater"`):

- $S_1(y)$ corresponds to **sensitivity** at threshold y
- $S_0(y)$ corresponds to the **false positive rate (FPR) = 1 - specificity**

For tests where smaller values indicate disease (`testdirection = "less"`):

- $S_1(y)$ corresponds to the **false negative rate (FNR) = 1 - sensitivity**
- $S_0(y)$ corresponds to **specificity**

## In a nutshell

By modelling the diagnostic marker as a latent continuous outcome, the Hoyer model:

- naturally accommodates **multiple thresholds per study**
- provides a unified framework for **interval-censored diagnostic data**
- preserves the correlation between sensitivity and specificity

The following parameters have to be estimated: $\beta_{0},\; \beta_{1},\; \lambda_0,\; \lambda_1,\; \sigma_0^2,\; \sigma_1^2,\; \text{and } \rho$, where $\lambda_0$ and $\lambda_1$ are transformations of $\phi_0$ and $\phi_1$.

To obtain initial values, it is necessary to specify a smallest (\<\< $\min y$) and largest (\>\> $\max y$) value across all studies, essentially replacing negative and positive infinity for the left and right censored intervals.

## How do I fit the Hoyer model?

```{r}
#| echo: TRUE
data("diabetes")
head(diabetes)
summary(diabetes$threshold)
```

The argument `eval_threshold` specifies the values at which sensitivities
and specificities are evaluated after model fitting.

```{r}
#| echo: TRUE
#| results: 'hide'
hoyer <- fitHoyer(data=diabetes,
                  TP=TP,
                  FP=FP,
                  FN=FN,
                  TN=TN,
                  threshold=threshold,
                  study=study,
                  dist="loglogistic",
                  testdirection="greater",
                  eval_threshold=c(5.0,5.5,6.0,6.5,7.0,7.5),
                  smallest=2,
                  largest=10)
```

```{r}
#| echo: TRUE
hoyer
summary(hoyer)
```

## How do we get a summary ROC plot?

```{r}
#| echo: TRUE
#| fig.height: 8
#| fig.width: 8
plot(hoyer,main="Diagnostic Test Accuracy Meta-Analysis\nLoglogistic Model")
plot(hoyer,size="sampsize",scale=0.025,main="Diagnostic Test Accuracy Meta-Analysis\nLoglogistic Model")
plot(hoyer,size="se",scale=0.025,main="Diagnostic Test Accuracy Meta-Analysis\nLoglogistic Model")
```

## How does `testdirection="less"` work?

We use the synthetic anaemia data set. Smaller values in haemoglobin indicate disease.

```{r}
#| echo: TRUE
data("anaemia")
summary(anaemia$threshold)
```
```{r}
#| echo: TRUE
#| results: 'hide'
less <-  fitHoyer(data=anaemia,
                  TP=TP,
                  FP=FP,
                  FN=FN,
                  TN=TN,
                  threshold=threshold,
                  study=study,
                  dist="loglogistic",
                  testdirection="less",
                  eval_threshold=c(8,9,10,11,12,13),
                  smallest=5,
                  largest=17)
```
```{r}
#| echo: TRUE
#| fig.height: 8
#| fig.width: 8
less
summary(less)
plot(less,main="Diagnostic Test Accuracy Meta-Analysis\nLoglogistic Model")
```

## How does `fitHoyer` work internally?

### Step 1: How do we restructure the data set?

```{r}
#| echo: TRUE
res <- restructure(
    data = diabetes,
    TP = TP,
    FP = FP,
    FN = FN,
    TN = TN,
    study = study,
    threshold = threshold,
    testdirection = "greater",
    smallest = 2,
    largest = 10)
res
```


### Step 2: How do we get initial values?

```{r}
#| echo: TRUE
init <- getInitParms(res$restructured,
                     dist="loglogistic")
init
```

### Step 3: How do we fit the model?

```{r}
#| echo: TRUE
#| results: 'hide'
hoyerAFT <- fitHoyerAFT(data=res,
                        init=init,
                        threshold=c(5.0,5.5,6.0,6.5,7.0,7.5))
```

```{r}
#| echo: TRUE
hoyerAFT
summary(hoyerAFT)
```


## References

Reitsma, J. B., et al. (2005).
Bivariate analysis of sensitivity and specificity produces informative summary measures in diagnostic reviews.
*Journal of Clinical Epidemiology*, 58(10), 982–990.

Rutter, C. M., & Gatsonis, C. A. (2001).
A hierarchical regression approach to meta-analysis of diagnostic test accuracy evaluations.
*Statistics in Medicine*, 20(19), 2865–2884.

Harbord, R. M., Deeks, J. J., Egger, M., Whiting, P., & Sterne, J. A. C. (2007).
A unification of models for meta-analysis of diagnostic accuracy studies.
*Biostatistics*, 8(2), 239–251.

Riley, R. D., Ensor, J., Jackson, D., & Burke, D. L. (2018).
Deriving percentage study weights in multi-parameter meta-analysis models.
*Statistical Methods in Medical Research*, 27(10), 2885–2905.

Hoyer, A., Hirt, S., Kuss, O. (2018).
Meta-analysis of full ROC curves using bivariate time-to-event models for interval-censored data.
*Research Synthesis Methods*, 9(1).

Deeks, J. J., Bossuyt, P. M., Leeflang, M. M., & Takwoingi, Y. (editors) (2023).
Cochrane Handbook for Systematic Reviews of Diagnostic Test Accuracy.
Version 2.0 (updated July 2023). Cochrane.
Available from <https://training.cochrane.org/handbook-diagnostic-test-accuracy/current>


